Programming in R - tutorial :  Matrix Data Type
========================================================
<i> Author: Abhinav Agrawal </i>

<b>Matrix</b> is a data type in R with the dimension attribute - the rows & the columns.  It has the elements of same class type. We can have character, integer or complex elements in the matrices and so on. We cannot have elements of mixed modes/class types such as both integer and character elements in the same matrix.<br>

Let us start with <b>creating matrix</b>
```{r}

m1 <- 1:4 # m1 is a vector object with a sequence of elements 
dim(m1) <- c(2,2) # define the dimension as 2X2 (2 rows and 2 columns) to make it a matrix
m1 # prints the matrix
class(m1) # show the class type, in this case it would be matrix class
typeof(m1) # mode of the elements. In this case elements stored in the matrix are integer type 
nrow(m1)  # outputs the number of rows in the matrix
ncol(m1)  # outputs the number of columns in the matrix

m2 <- matrix(nrow = 2, ncol = 2) # we will get the matrix filled with NA as we have not defined the matrix values
m2
dim(m2)
attributes(m2)

m2 <- matrix(seq(1,4,1), nrow = 2, ncol = 2) # used a seq() to get a sequence of elements
m2 # note that by default the matrix is filled column wise. To fill row wise, we need to tell R exclusively. 

m2 <- matrix(seq(1,4,1), nrow = 2, ncol = 2, byrow = T) # byrow = T would tell R to fill the elements row wise
m2 

matrix(c(1,2,3,4), c(2,2)) # elements and dimensions of the matrix both defined using a c()function

m2 <- matrix(c(0+1i,1+2i,3+4i,4+5i), nrow = 2, ncol = 2, byrow = T) 
# byrow = T would tell R to fill the elements row wise
m2 
class(m2)
typeof(m2)

m2 <- matrix(c("A", "B", "C", "D"), nrow = 2, ncol = 2, byrow = T) 
# byrow = T would tell R to fill the elements row wise
m2 
class(m2)
typeof(m2)
```
Now, let us do some <b>subsetting</b> in matrix
```{r}
m3 <- matrix(1:9, nrow = 3, ncol = 3)
m3
m3[1,1] # extracting an element by giving its location
m3[,1] # extracting row elements of a specific column, in this case column 1
m3[,2] # extracting row elements of a specific column, in this case column 2
m3[,3] # extracting row elements of a specific column, in this case column 3
m3[1,] # extracting column elements of a specific row, in this case row 1
m3[2,] # extracting column elements of a specific row, in this case row 2
m3[3,] # extracting column elements of a specific row, in this case row 3
diag(m3) # extracts the diagonal elements of the matrix
```

<b> Assign row names and column names </b> to the matrix
```{r}
m4 <- matrix(1:4, byrow = T, ncol=2)
m4

colnames(m4) <- c("COL1", "COL2")
rownames(m4) <- c("ROW1", "ROW2")
m4

dimnames(m4) <- list(c("COL1", "COL2"),c("ROW1", "ROW2"))
m4
```

Using <b>cbind()</b> and <b>rbind()</b> functions in the matrix:
```{r}
m6 <- 0:3
m7 <- 2:5 
# both the vectors should be of either the same length or length of bigger vector should be the multiple of shorter length vector

cbind(m6,m7) # fill column wise
rbind(m6,m7) # fill row wise

```

<b> Mathematical operations </b> on matrix

```{r}
# create 2 matrices
mat10 <- matrix(5:8, c(2,2))
mat11 <- matrix(seq(1,8,2), c(2,2))

mat10 + 10 # Addition of scalar to a matrix

mat10 + mat11 # Adding two matrices

mat10 - mat11 # Substraction of 2 matrices

mat10 * 2 # Matrix multiplication (scalar multiplied to a matrix)

mat10 * mat11 # Multiplication of 2 matrices (element wise multiplication)

mat10 %*% mat11 # Multiplication of matrix (the actual matrix multiplication)

t(mat11) # Transpose of a matrix

solve(mat10) # Inverse of a mtarix

rowSums(mat10) # sum of rows

colSums(mat10) # sum of columns

colMeans(mat10) # return the means of the columns 

rowMeans(mat10) # # return the means of the columns 

sum(mat10) # sum of all the elements in the matrix

eigen(mat10) # to calculate the eigen values

```
We may want to convert a vector to a matrix - using the <b> as.matrix() </b> function for exclusive coercion
```{r}
mn <- 1:6
mn # mn is a vector object

as.matrix(mn) # as.matrix() function converts the vector mn to a matrix object

```


